---
title: 
author: 
date:
output:
  bookdown::html_book:
    toc: yes
    css: toc.css
---
```{r setupweek7, include=FALSE}
knitr::opts_chunk$set(fig.width = 8, collapse = TRUE)
knitr::opts_chunk$set(echo = TRUE)
```

# Week 7 For loops and the apply family of functions

A few useful commands: function(), is.na, which, var, length, for(){ }, 
 points, print, paste, plot, unique, sample
 
for loops: 
In many languages, the best way to repeat a calculation is to use a for-loop:
For example, we could square each number 1 to 10
```{r, echo=T}
squares = rep(NA, 10) # use rep to create a vector length 10 of NAs to store the result
for (i in 1:10) { # for loop
  squares[i] = i^2
}
squares
```
An alternative to for-loops in R is using the 'apply' family, while for-loops apply a function to one item at a time and then go on to the next one, "apply" applies functions to every item at once
## apply family
### sapply
There are several apply functions that vary in the output the return and vary somewhat in the input they require. We'll go over **sapply** "simplifying" apply which returns a vector, first.
```{r, echo=T}
#?sapply 
# syntax: sapply(X = object_to_repeat_over, FUN = function_to_repeat)
# simple example of sapply over a vector
# we can use an in-line function definition
sapply(1:10, function(x)  x^2)
# equivalently, we can define our own functions separately for sapply
# e.g. a function that calculates the area of a circle radius r, pi*r^2
areaCircle = function(r){
  return(pi * r^2)
}
sapply(1:10, areaCircle)
# in R, we can also just use short-hand for simple vector calculations:
pi*(1:10)^2
# but unlike the short-hand, sapply can also iterate over elements in a list
listy = list(a = TRUE, b = c("a", "b", "c"), c = 10:100)
str(listy) # look at the structure of 'listy'
length(listy) # look at the length of 'listy'
# use sapply to return a vector for length of each object within the list
sapply(listy, FUN = length) 
```
You can also use sapply to create plots! For example, use sapply to plot these 4 dataframes at once:
```{r, label='5-33', echo=T}
df1 = data.frame(x1 = 1:10, y1 = 1:10)
df2 = data.frame(x2 = 1:10, y2 = -1:-10)
df3 = data.frame(x3 = 1:10, y3 = 10:1)
df4 = data.frame(x4 = 1:10, y4 = 1:10)
my_list = list(df1, df2, df3, df4) # put 4 data frames together in a list
par(mfrow = c(2,2)) # set up frame for 4 plots
sapply(my_list, plot) # plot my_list with sapply
```
### apply 
The apply function is highly useful for applying a function to rows or columns of a dataframe or matrix. 
Example syntax for the dataframe or matrix X:
`apply(X = over this object, MARGIN 1 for rows or 2 for columns,FUN = apply this function)`
You can also use apply on a dataframe we worked with earlier the states data to plot each column against Population
```{r, echo=T}
#load in the data
data(state)
states = as.data.frame(state.x77) # convert data to a familiar format - data frame
str(states) # let's take a look at the dataframe
# calculate the mean for each column
apply(states, 2, mean)
# note you could get this with colMeans() or summary(), along with the min and max and other values, but there may be instances where you only want the mean
# you could also plot each column against Population in ggplot
```
```{r, label='5-29', echo=T}
apply(states, 2, FUN = function(i) ggplot(states, aes(x=Population, y = i))+geom_point()+geom_smooth(method="lm")+theme_classic())
```
We can do the same things across all rows. But if you want to plot all the rows as we did the columns above, I suggest you do that with a smaller dataset than the states dataframe.
```{r, echo=T} 
#calculate the sum across each row in states
apply(states, 1, sum)
```
### lapply -- "list" apply
We'll just show a quick example of lapply. It works in the same way as sapply, but returns a list instead of a vector.
```{r, echo=T}
lapply(1:10, function(x)  x^2) # lapply returns list
sapply(1:10, function(x)  x^2, simplify = FALSE) # same as an lapply
sapply(1:10, function(x)  x^2) # default is simplify = TRUE which retuns a vector
```
### tapply - "per Type" apply 
The tapply function is one of my favorites because it is a really great way to sumarize data that has multiple categorical variables that can be 

```{r, echo=T}
# load state data again, you can skip this if you already have it loaded
data(state)
states = as.data.frame(state.x77) # convert data to a familiar format - data frame
str(states) # let's take a look at the dataframe
# example syntax --- tapply(variable of interest, grouping variable, function)
# for each US region in our dataset, finds the mean of Frost for states in that region
tapply(states$Frost, state.region, mean) # state.region contains the region information for each state
# you can nest apply statements! Let's find the region average for all the variables in the states dataset
apply(states,
      2, # apply over columns of my_states
      function(x) tapply(x, state.region, mean)) # each column = variable of interest for tapply
```
## Exercise 2.3 apply and tapply
> # Exercise 2.3
A few useful commands: function(){ }, apply(), tapply(), hist(), dim(), prod(), sd()
> 1. what is the average population, income, and area of all 50 states ins the `states` dataset

<details><summary><span style="color: red;">Solution</span></summary>
<p>

```{r, echo=T}
# load state data
#?state
data(state)
# this data is stored in a slightly different way than other datasets we've used so far
states = as.data.frame(state.x77) # run this line of code to avoid later confusion
apply(states,2,mean)
#or an alternative that will get you only the columns requested
colMeans(states[,c("Population", "Income", "Area")])
```

</p>
</details>

> 2. what is the average area of the states from different regions of the country?
Hint: use the object state.region in your environment 

<details><summary><span style="color: red;">Solution</span></summary>
<p>
```{r, echo=T}
tapply(states$Area, state.region, mean)
```

<p>
</details>

> 3. Plot a histogram for each column in the states data (Population, Income, Illiteracy etc.)

<details><summary><span style="color: red;">Solution</span></summary>
<p>

```{r, label='5-32', echo=T}
#how many columns do we have? 
dim(states)
par(mfrow = c(2,2)) # make your plot window show 2 rows and 2 columns at once
apply(states, 2, hist)
```

</p>
</details>

> 4. let's assume that we don't want to live in a state with high illiteracy, high murder, and many freezing days; also assume that each of these factors contribute equally to our opinion (Illiteracy * Murder * Frost) = undesirable
  What 10 states should we avoid? # hint use prod(); and maybe order()
  
<details><summary><span style="color: red;">Solution</span></summary>
<p>

```{r, echo=T}
livability <- apply(states[,c("Illiteracy", "Murder", "Frost")], 1, prod) # subset to variables of interest
livability[order(livability, decreasing = T)][1:10] # top ten least livable states
```

</p>
</details>

> 5. use sapply() to plot a histogram of the data below 4 times, in 4 different colors. 
  For extra style, title the plot by it's color, e.g. the red plot is titled "red"
  
```{r, echo=T}
data_to_plot = c(1,3,4,5,6,3,3,4,5,1,1,1,1,1)
par(mfrow = c(2,2))# run this line to set your plot to make 4 plots in total (2rows, 2columns)
```

<details><summary><span style="color: red;">Solution</span></summary>
<p>

```{r, label='5-30', echo=T}
data_to_plot = c(1,3,4,5,6,3,3,4,5,1,1,1,1,1)
my_colors = c("deeppink", "red", "blue", "darkgreen")
par(mfrow = c(2,2)) # extra styling, plots in a 2x2 grid
sapply(my_colors, FUN = function(i) hist(data_to_plot, main = i, col = i))
```

</p>
</details>

> 6. Standardize all the variables in the states dataset and save your answer to a new dataframe, states_standardized
Hint: to standardize a variable, you subtract the mean and divide by the standard deviation (sd)

<details><summary><span style="color: red;">Solution</span></summary>
<p>

```{r, echo=T}
states_standardized = apply(states, 2, function(x) (x-mean(x))/sd(x))
# original:
head(states)
# standardized
head(states_standardized)
```

</p>
</details>

> 7. Create a histogram again for each variable in the states data, but this time label each histogram with the variable names when you plot
Hint: instead of using apply to iterate over the columns themselves, you can often iterate over the column names with sapply

<details><summary><span style="color: red;">Solution</span></summary>
<p>
```{r, label='5-31', echo=T}
par(mfrow = c(2,2))
sapply(colnames(states), function(x)hist(states[ , x],main = x, xlab = x,col = "darkblue"))
```
</p>
</details>
