---
title: 
author: 
date: 
output:
  bookdown::html_book:
    toc: yes
    css: toc.css
---

```{r setupweek11, include=FALSE}
knitr::opts_chunk$set(fig.width = 8, echo=F, error=T, message=F)
knitr::opts_knit$set(root.dir = "C:/Users/SAPCaps/MarineGenomicsSemester/data/Week11_gea/")
```

# Genome Environment Association

The lecture for this week can be found [here](https://github.com/BayLab/MarineGenomicsSemester/blob/main/ppt/MarineGenomics_Lecture_w11.pdf).

All of the code presented in this weeks class comes modified from a former Marine Genomics student and recent UC Davis graduate, Camille Rumberger! Thank you Camille!

## Download the data

For this week we are again using data from the wonderful Xuereb et al. 2018 paper [here](https://onlinelibrary.wiley.com/doi/abs/10.1111/mec.14589?casa_token=aBVeQUaZ6UEAAAAA:vt9cQFbQ-65F1erD-1Uq0DAWEaZ75fDhepkPomc4RMPAqQgntIcm0btk842SxvaraM2VdlZ5nwoHUhCy)
And consistes of a vcf file with 3966 Snps from 685 pacific sea cucumbers. 

```html

wget https://raw.githubusercontent.com/BayLab/MarineGenomicsData/main/week11_semester.tar.gz
tar -xzvf week11_semester.tar.gz

```


## install R packages


```{r echo=T, message=FALSE}

#devtools::install_github("bcm-uga/lfmm")


# get packages
#install.packages(c('lfmm','psych','vegan','ggplot2','data.table','sdmpredictors','rgdal','raster'), dependencies = T)
#install.packages("BiocManager")
#BiocManager::install("LEA", force=T)

# call programs
library(lfmm)
library(psych)
library(vegan)
library(LEA)
library(data.table)
library(sdmpredictors)
library(leaflet)
library(ggplot2)
library(rgdal)
library(raster)

```


## Get the Environmental Data


now we need to download the environmental data for our study. We're just going to look at a few variables but there are many others that you could choose from here: https://www.bio-oracle.org/explore-data.php

This step is relatively time consuming so while the code is provided below, in class we'll be working with a file that we previously downloaded


```{r, echo=T, warning=F}

environ <- load_layers(layercodes = c("BO_chlomean","BO_ph", "BO_salinity","BO_sstmax","BO_sstmean"))

#store these as objects that we can use later

chlomean<-load_layers("BO_chlomean")
ph<-load_layers("BO_ph")
salinity<-load_layers("BO_salinity")
sst_max<-load_layers("BO_sstmax")
sst_mean<-load_layers("BO_sstmean")


```


Now we'll read in our metadata, which has lat and lon for each sample


```{r, echo=T}
meta<-read.csv("californicus_metadata.csv")

#make a new dataframe that just contains the lat and lon and label them "lat" and "long"
#notice that the original metafile mislabels lat and lon, oops!

sites<-cbind("lat" =  meta$LONG, "lon" = meta$LAT)

head(sites)

```

## Extract Site Specific Information from our Envirornmental Layers


```{r echo=T}

sites_environ <- data.frame(Name=sites, extract(environ,sites))
head(sites_environ)

```

That produces for us a site specific environment file. We now need to convert this file into a format that we can save it as a matrix and export it as an environment file (i.e., in the format that the next step need it to be in)

```{r, echo=T}

#remove lat and lon and convert to matrix

sites_environ_matrix<-as.matrix(sites_environ[,-c(1,2)])

#remove any Na's

sites_environ_matrix_nas<-na.omit(sites_environ_matrix)

#write the file to an env file

write.env(sites_environ_matrix_nas, "sites_environ_matrix.env")


```
## make a map of our environmental data

```{r, label='geamap', echo=T}
library(sdmpredictors)

#download a raster file for env variable that we want to look at

#what are are min amd max lat and lon

range(sites_environ$Name.lon)

#crop file to fit the area we want
ne.pacific<-extent(-140, -120, 40, 60)

sst_max.crop<-crop(sst_max, ne.pacific)


#make a nice color ramp and plot the map

my.colors = colorRampPalette(c("#5E85B8","#EDF0C0","#C13127"))
plot(sst_max.crop,col=my.colors(1000),axes=FALSE, box=FALSE)
title(cex.sub = 1.25, sub = "Maximum temperature at the sea surface (ºC)")


```




## Load in genetic data


Now we need to load in our genetic data



```{r, echo=T}

#convert our vcf file to lfmm

vcf2lfmm('filtered_3699snps_californicus_685inds.recode.vcf', 'filtered_3699snps_californicus.lfmm')

sea_cuc_lfmm<-read.lfmm('filtered_3699snps_californicus.lfmm')

#and convert to geno

lfmm2geno('filtered_3699snps_californicus.lfmm', 'filtered_3699snps_californicus.geno')

#read in geno file

sea_cuc_geno<-read.geno('filtered_3699snps_californicus.geno')




```



```{r, label='gea1', echo=T, message=F}


# create a snmf object

sea_cuc200.snmf <- snmf("filtered_3699snps_californicus.geno", K = 1:10, entropy=T, ploidy = 2, project = "new")
project=load.snmfProject("filtered_3699snps_californicus.snmfProject")

#plot values of cross-entropy criteron of k
plot(sea_cuc200.snmf)

```
```{r, echo=F}

# Genome scan for selection using environmental variables 
#remove.lfmmProject('Spurp.prelim.snps24.lfmmProject')
#sea_cuc.lfmm <- lfmm("data/Week11_gea/filtered_3699snps_californicus_685inds.recode.lfmm", "data/Week11_gea/sites_environ_matrix.env", K=1, rep=1, project = "new")

# load project
project = load.lfmmProject("data/Week11_gea/filtered_3699snps_californicus_685inds.lfmmProject")

```

Now we can check and see which of our environmental variables is association with genomic variation

```{r, echo=T}
#get the z scores from each run

zs1 <- z.scores(sea_cuc.lfmm, K=1, d=1)
zs2 <- z.scores(sea_cuc.lfmm, K=1, d=2)
zs3 <- z.scores(sea_cuc.lfmm, K=1, d=3)
zs4 <- z.scores(sea_cuc.lfmm, K=1, d=4)
zs5 <- z.scores(sea_cuc.lfmm, K=1, d=5)

# compute the genomic inflation factor lambda

lambda1=median(zs1^2)/qchisq(0.5, df=1)
lambda2=median(zs2^2)/qchisq(0.5, df=1)
lambda3=median(zs3^2)/qchisq(0.5, df=1)
lambda4=median(zs4^2)/qchisq(0.5, df=1)
lambda5=median(zs5^2)/qchisq(0.5, df=1)

#then compute adjusted p-values

adj.p.val1<-pchisq(zs1^2/lambda1, df=1, lower=F)
adj.p.val2<-pchisq(zs2^2/lambda2, df=1, lower=F)
adj.p.val3<-pchisq(zs3^2/lambda3, df=1, lower=F)
adj.p.val4<-pchisq(zs4^2/lambda4, df=1, lower=F)
adj.p.val5<-pchisq(zs5^2/lambda5, df=1, lower=F)

#then plot that p-value

hist(adj.p.val1, col="blue")
hist(adj.p.val2, col="blue")
hist(adj.p.val3, col="blue")
hist(adj.p.val4, col="blue")
hist(adj.p.val5, col="blue")


# control of false discoveries
# to correct for multiple testing, we can apply the Benjamini-Hochberg algorithm
# L is number of loci
L=3699
#fdr level q
q = 0.1
w = which(sort(adj.p.val4) < q * (1:L)/L)
# candidates are then
candidates.4 = order(adj.p.val4)[w]

length(candidates.4)


```






## Exercises


















