---
title: 
author: 
date: 
output:
  bookdown::html_book:
    toc: yes
    css: toc.css
---

```{r setupweek12, include=FALSE}
knitr::opts_chunk$set(echo = F,error = F,fig.width = 8,message = FALSE)
knitr::opts_knit$set(root.dir = "C:/Users/SAPCaps/MarineGenomicsSemester/data/Week12_structure/")
```

# Population Structure via conStruct

For this week we will be exploring another way to document population genetic structure (the first way we covered was PCA in Week 8) via a "Structure" plot implemented in the R package conStruct.

conStruct itself has a nice series of tutorials with example data that you can find [here](https://cran.r-project.org/web/packages/conStruct/vignettes/format-data.html)

We'll be using data from the [Xuereb et al. paper on P. californicus](https://onlinelibrary.wiley.com/doi/abs/10.1111/mec.14589). This data consists of a vcf file that is in structure format and contains SNP data (3699 SNPs) from 717 individuals (this is the same data we used for our PCA plot in week 9 for pcadapt, and for week 11 for GEA). 

The lecture for this week can be found [here](https://github.com/BayLab/MarineGenomics/blob/4f22a87ad800f98a6e138a166b8e94d016310914/ppt/Week9.pdf) and describes the basics of a structure plot.

## Download the data

We'll be using data from the [Xuereb et al. paper on P. californicus](https://onlinelibrary.wiley.com/doi/abs/10.1111/mec.14589). This data consists of a vcf file that is in structure format and contains SNP data (3699 SNPs) from 100 individuals (this is the same data we used for our PCA plot in week 9 for pcadapt, and for week 11 for GEA, just with fewer individuals for this week). 

We'll download a metafile as well that has sample coordinates and names

```html

wget https://raw.githubusercontent.com/BayLab/MarineGenomicsData/main/week12_semester.tar.gz

tar -xzvf week12_semester.tar.gz

```
## install conStruct in R

```{r, echo=F, message=FALSE, warning=F, results=F}

#install.packages("conStruct")

#load the package

library(conStruct)


```

## read in the data

This first step reads in our file which is in structure format and converts it to conStruct format (there are lots of file formats in population genetics, isn't it fun?)

We need to make sure that we have the file coding right and this can be confirmed by looking at the structure file in bash to make sure that our samples start at the 3rd row and 3rd column and that our missing data is encoded as a -9.

You can do the above by using wc -l to find the number of lines in the structure file. 

```{r, echo=T}

conStruct.data <- structure2conStruct(infile="filtered_3699snps_100indcalifornicus.recode.strct_in",
                                      onerowperind = TRUE,
                                      start.loci = 3,
                                      start.samples = 3,
                                      missing.datum = -9,
                                      outfile = "Construct_100inds")

```

## Get our meta data read into R and ready to use

This file contains the lat and lon of our samples as well as the site name and individual id.

For conStruct we can make an object that contains the lat and lon and has the site name as the row names.


```{r, echo=T}

#read in the data
meta<-read.table("californicus_metadata.csv", header=T, sep=",")

#pull out the lat and lon and put it in a new matrix object

sites<-as.matrix(meta[-c(51:667), c("LAT", "LONG")])

#assign row names that are the site id for each sample

rownames(sites)<-meta$SITE[-c(51:667)]


```


## Run a structure analysis

We now have our data read into conStruct and converted to conStruct format and our site information from the metadata in a matrix format.

We can now run a structure analyses. For these analyses we chose a value of K (the number of clusters) and typically run the analyses for a range of K (2:7 for example).

We'll run K = 2 through K = 4 and then we'll do some model comparison to discuss whether one of these is "better" than the other. 

And we'll only run it with 1 chains and 10 iterations. This is simply so that the program runs quickly within our class time. For a "real" analysis you'd want to run more than one chain (2-5) and iterations up to 1000. It's important to compare the different chains to ensure the analyses converges, as described [here](https://cran.r-project.org/web/packages/conStruct/vignettes/run-conStruct.html#model-diagnosis)

```{r, echo=T, message=F, warning=F, results=F}


my.run <- conStruct(spatial = FALSE,
                    K = 1,
                    freqs = conStruct.data,
                    geoDist = NULL,
                    coords = sites,
                    prefix = "Sea_cucu_K1",
                    n.chains = 1,
                    n.iter = 10, 
                    make.figs=F)

my.run <- conStruct(spatial = FALSE,
                    K = 2,
                    freqs = conStruct.data,
                    geoDist = NULL,
                    coords = sites,
                    prefix = "Sea_cucu_K2",
                    n.chains = 1,
                    n.iter = 10, 
                    make.figs=F)



my.run <- conStruct(spatial = FALSE,
                    K = 3,
                    freqs = conStruct.data,
                    geoDist = NULL,
                    coords = sites,
                    prefix = "Sea_cucu_K3",
                    n.chains = 1,
                    n.iter = 10, 
                    make.figs=F)


my.run <- conStruct(spatial = FALSE,
                    K = 4,
                    freqs = conStruct.data,
                    geoDist = NULL,
                    coords = sites,
                    prefix = "Sea_cucu_K4",
                    n.chains = 1,
                    n.iter = 10, 
                    make.figs=F)



my.run <- conStruct(spatial = FALSE,
                    K = 5,
                    freqs = conStruct.data,
                    geoDist = NULL,
                    coords = sites,
                    prefix = "Sea_cucu_K5",
                    n.chains = 1,
                    n.iter = 10, 
                    make.figs=F)



```

This step produces several output files (a conStruct.results file and a data.block file) called after whatever you put in the prefix = section (sea_cucu_k2 in our case). We need to load that output file back into R to access all the things it produced.

Let's just look at the first one for k = 2.

```{r, echo=T}

#load the construct results file
load("Sea_cucu_K2_conStruct.results.Robj")

admix.props<- conStruct.results$chain_1$MAP$admix.proportions

#load the data block file
load("Sea_cucu_K2_data.block.Robj")


# make a structure plot
make.structure.plot(admix.proportions = admix.props,
                    mar = c(4,4,2,2))

```


## Comparisons across different values of K

You may want to show a single structure plot for one value of K and may want that one plot to be the "best" fitting model for your data. 

There are a few ways to do this, however often there's not a single "best" model and care should be taken as models with more parameters often appear to be a better fit, but are not necessarily more biologically meaningful.

We'll be looking at how different levels of K contribute to describing patterns of variation in our data. Code taken and modified from the vignette [here](https://cran.r-project.org/web/packages/conStruct/vignettes/model-comparison.html)

The idea here is that if we can visualize how much different levels of K contribute to describing our data we can tell whether we're getting more information from a K = 5 compared to a K = 2 run. 


```{r, echo=T}

layer.contributions <- matrix(NA,nrow=5,ncol=5)

# load the conStruct.results.Robj and data.block.Robj
#   files saved at the end of a conStruct run
load("Sea_cucu_K1_conStruct.results.Robj")
load("Sea_cucu_K1_data.block.Robj")

# calculate layer contributions
layer.contributions[,1] <- c(calculate.layer.contribution(conStruct.results[[1]],data.block), rep(0,4))
tmp <- conStruct.results[[1]]$MAP$admix.proportions

for(i in 2:5){
    # load the conStruct.results.Robj and data.block.Robj
    #   files saved at the end of a conStruct run
    load(sprintf("Sea_cucu_K%s_conStruct.results.Robj",i))
    load(sprintf("Sea_cucu_K%s_data.block.Robj",i))
    
    # match layers up across runs to keep plotting colors consistent
    #   for the same layers in different runs
    tmp.order <- match.layers.x.runs(tmp,conStruct.results[[1]]$MAP$admix.proportions)  

    # calculate layer contributions
    layer.contributions[,i] <- c(calculate.layer.contribution(conStruct.results=conStruct.results[[1]],
                                                             data.block=data.block,
                                                             layer.order=tmp.order),
                                    rep(0,5-i))
    tmp <- conStruct.results[[1]]$MAP$admix.proportions[,tmp.order]
}

```


We'll then plot these results as a barplot

```{r, echo=T}


barplot(layer.contributions,
        col=c("blue", "red", "goldenrod1", "forestgreen", "darkorchid1"),
        xlab="",
        ylab="layer contributions",
        names.arg=paste0("K=",1:5))


```

## Exercises



1. change the colors in the plot

2. order the plot by latitude



